Trending assessment for nowcasts, measurements, and forecasts

### 2024-03-04 sections/03-trending.tex
Update on Overleaf.

As outlined in Section~\ref{sec:introduction}, we refer by trending to the consistency of an observed change with the change issued by another measurement, a nowcast, or a forecast.
In this section, we want to specify trending and lay out different methods of measuring and visualizing trending. 
In general, trending can be assessed for different lags $l$ separately, and the relevance of the considered lags has to be determined based on the question at hand.
Throughout the section, we call a measurement, nowcast, or forecast a \textit{prediction} or \textit{signal} whenever the application area is insignificant.
Before introducing measures for assessing trending, we give remarks on computing the predicted change and notation in Section~\ref{subsec:notation}.
Section~\ref{subsec:trending-basics} introduces trending basics and visualizes trending in a synthetical example.
In Section~\ref{subsec:trending-measures}, various quantification methods from different disciplines and tasks are examined.
The approaches are extended to the trending setting in Section~\ref{subsec:trending-noise}.
Section~\ref{subsec:trending-cond-prob} introduces a new graphical method of trending assessment and briefly introduces bootstrap methods for computing confidence intervals for the measures.

@@

As noted above, trending refers to observed and predicted changes for a time lag $l$.
The observed change is straightforward to compute for measurements, nowcasts, and forecasts.
Let $\mathbf{y} = (y_t)_{t=0}^T$ denote the true values or gold standard measurements up to time $T$.
The sequence of observed change is then given by the differences of values in $\mathbf{y}$ with lag $l$, that is
\begin{equation}\label{eq:diffy}
    \diffylt = (y_{t+l} - y_t) \quad \text{for}\ t = 1, \dots, T - l.
\end{equation}
As noted in the introduction, the definition of predicted change depends on the context.
Conveying the computation for $y$ directly, the signal's predicted change would be the difference between the signal at time $t + l$ and the true value at $t$.
This approach lags in three practical aspects.
For nowcasting and forecasting, the true value at time $t$ might not be known with a considerable time lag.
In the example in Section~\ref{sec:application-covid}, for example, the time lag is more than 80 days.
Until the true values are publicly available, the change cannot be computed, and no actions can be taken based on it.
Thus, it is more practically relevant to use updated signal values for time $t$ as the reference value in this setting.

@@

Taking the three points together, the signal's predicted change is based on the type of signal.
To account for several signals for a forecast time $t$ at different issue times $\tau$, we denote those by $x_{t | \tau}$.
For nowcasting, $t$ and $\tau$ are usually close and possibly negative; for forecasting, $t$ is usually larger than $\tau$.
For measurements, this distinction is not necessary.
Table~\ref{tab:notation} summarizes the notation for the different signal types and the predicted change computation.

@@

Throughout the remainder of the section, we use the notation of differences, $\diffxl$ and $\diffyl$.
While the computation of $\diffx$ differs in different contexts, the trending assessment is similar.
In the following, we omit the difference's lag $l$ for the ease of notation; $\diffx$ and $\diffy$ refer to $\diffxl$ and $\diffyl$ for a common lag $l$.
As noted above, a signal is trending perfectly if all occurred directions of change are predicted correctly; that is, the sign of all elements of $\diffx$ and $\diffy$ coincide.
Perfect trending is hardly the case in practice, but a weaker requirement is needed.
Accordingly, when assessing the trending, we examine the statistical consistency of $\sign(\diffx)$ and $\sign(\diffy)$.
A simple yet insightful method is the four-quadrant plot.
For example, it is known in cardiac output measurement analysis~\parencite{Saugel2015,perrino1998intraoperative}. 
Thereby, the occurred changes are plotted together with the predicted changes, that is, $(\diffy_t, \diffx_t)$ for $t = 1, \dots, T$.
Thus, the x-axis of a four-quadrant plot shows the true value differences, whereas the y-axis displays prediction data differences.
Points in the green upper right and lower left quadrants reflect a correct trending for the respective time step, whereas points in the remaining red quadrants show incorrect predicted changes.
Figure~\ref{fig:trending_basic_4q} displays a basic four-quadrant plot.
Points 2, 3, 5, and 6 show trending, whereas points 1, 4, and 7 count for anti-trending behavior.
Figure~\ref{fig:trending_basic_4q_sample} shows a four-quadrant plot for a more extensive simulated data set with $T=1461$, for example, four years of daily data.
The data generation is described in Appendix~\ref{subsec:app-trending-data-generation} and will carry through this section.
The four-quadrant plot resembles a butterfly-like shape, with more data in the upper right and lower left quadrants than in the remaining quadrants.
The four-quadrant plot is intuitive to interpret, and the magnitude and direction of change are shown simultaneously.
It can be extended by including information on the date of difference in the point color.
In Figure~\ref{fig:trending_basic_4q_sample_color}, blue points refer to earlier measurements and turn green for larger $t$.
The lag is always one time step.
However, four-quadrant plots become crowded for larger datasets, and the sequential information on the differences is complex to assess thoroughly.
There are other visualization techniques, such as polar plots or the Bland-Altmann analysis.
However, they lack the clearness and intuition of the four-quadrant plot while adding no more information on the trending~\parencite{Saugel2015}.

\begin{figure}
\centering
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_without_excl}
\caption{Basic four-quadrant plot.} \label{fig:trending_basic_4q}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_box}
\caption{Four-quadrant plot with rectangle exclusion area.}\label{fig:trending_basic_4q_excl_box}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_axis}
\caption{Four-quadrant plot with horizontal exclusion area.} \label{fig:trending_basic_4q_excl_axis}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_cross}
\caption{Four-quadrant plot with crossed exclusion area.}\label{fig:trending_basic_4q_excl_cross}
\end{subfigure}%
\caption{Illustrations of the four-quadrant plot with sample points and with and without exclusion areas. }
\label{fig:trending_4q}
\end{figure}

>>>>

As outlined in Section~\ref{sec:introduction}, we refer by trending to the consistency of an observed change with the change issued by another measurement, a nowcast, or a forecast.
In this section, we want to specify trending and lay out different methods of measuring and visualizing trending. 
In general, trending can be assessed for different lags $l$ separately, and the relevance of the considered lags has to be determined based on the question at hand.
Throughout the section, we call a measurement, nowcast, or forecast a \textit{prediction} or \textit{signal} whenever the application area is insignificant.
Before introducing measures for assessing trending, we give remarks on computing the predicted change and notation in Section~\ref{subsec:notation}.
Section~\ref{subsec:trending-basics} introduces trending basics and visualizes trending in a synthetical example.
In Section~\ref{subsec:trending-measures}, various methods from other disciplines and tasks are applied and examined.
The approaches are extended to account for noise and non-systematic effects in Section~\ref{subsec:trending-noise}.
Section~\ref{subsec:trending-cond-prob} introduces a new graphical method for local trending assessment and briefly introduces bootstrap methods for computing confidence intervals for the measures of the prior sections.
Trending assessment for probabilistic nowcasts and forecasts is treated in Section~\ref{subsec:probabilistic}.

@@

As noted above, trending refers to observed and predicted changes for a time lag $l$.
The \textit{observed} change is straightforward to compute for all types of signal.
Let $\mathbf{y} = (y_t)_{t=0}^T$ denote the true values or gold standard measurements up to time $T$.
The sequence of observed change is then given by the differences of values in $\mathbf{y}$ with lag $l$, that is
\begin{equation}\label{eq:diffy}
    \diffylt = (y_{t+l} - y_t) \quad \text{for}\ t = 1, \dots, T - l.
\end{equation}

As noted in the introduction, the definition of the predicted change depends on the context.
Applying the computation for $y$ directly, the signal's predicted change would be the difference between the signal at time $t + l$ and the true value at $t$.
This approach lags in three practical aspects.
For nowcasting and forecasting, the true value at time $t$ might not be known with a considerable time delay.
In the nowcasting application of Section~\ref{sec:application-covid}, for example, the time lag is more than 80 days.
Until the true values are publicly available, the change could not be computed, and no actions could be taken based on it.
Thus, it is more practically relevant to use updated signal values for time $t$ as the reference value in this setting.

@@

Taking the three points together, the signal's predicted change is based on the type of signal.
To account for several signals for a target time $t$ at different issue times $\tau$, we denote those by $x_{t | \tau}$.
For nowcasting, $\tau$ is usually not considerably larger than $t$; while $t$ is usually at least $\tau$ in forecasting.
For measurements, this distinction is not necessary.
Table~\ref{tab:notation} summarizes the notation for the different signal types and the predicted change computation.

@@

While the computation of $\diffxl$ differs in different contexts, the trending assessment is similar.
In the following, we omit the difference's lag $l$ for the ease of notation; $\diffx$ and $\diffy$ refer to $\diffxl$ and $\diffyl$ for a common lag $l$ as defined in Table~\ref{tab:notation}.
As noted above, a signal is trending perfectly if all occurred directions of change are predicted correctly; that is, the sign of all elements of $\diffx$ and $\diffy$ coincide.
Perfect trending is hardly the case in practice, but a weaker requirement is needed.
Accordingly, when assessing the trending, we examine the statistical consistency of $\sign(\diffx)$ and $\sign(\diffy)$.
A simple yet insightful method is the four-quadrant plot.
For example, it is known in cardiac output measurement analysis~\parencite{Saugel2015,perrino1998intraoperative}. 
Thereby, the occurred changes are plotted together with the predicted changes, that is, $(\diffy_t, \diffx_t)$ for $t = 1, \dots, T$.
Thus, the x-axis of a four-quadrant plot shows the true value differences, whereas the y-axis displays prediction data differences.
Points in the green upper right and lower left quadrants reflect a correct trending for the respective time step, whereas points in the remaining red quadrants show incorrect predicted changes.
Figure~\ref{fig:trending_basic_4q} displays a basic four-quadrant plot.
Points 2, 3, 5, and 6 show trending, whereas points 1, 4, and 7 count for anti-trending behavior.
Figure~\ref{fig:trending_basic_4q_sample} shows a four-quadrant plot for a more extensive simulated data set with $T=1461$, for example, four years of daily data.
The data generation is described in Appendix~\ref{subsec:app-trending-data-generation} and will carry through this section.
The four-quadrant plot resembles a butterfly-like shape, with more data in the upper right and lower left quadrants than in the remaining quadrants.
The four-quadrant plot is intuitive to interpret, and the magnitude and direction of change are shown simultaneously.
It can be extended by including information on the date of difference in the point color.
In Figure~\ref{fig:trending_basic_4q_sample_color}, blue points refer to earlier measurements and turn green for larger $t$.
The lag is always one time step.
However, four-quadrant plots become crowded for larger datasets, and the sequential information on the differences is complex to assess thoroughly.
There are other visualization techniques, such as polar plots or the Bland-Altmann analysis.
However, they lack the clearness and intuition of the four-quadrant plot while adding no more information on the trending~\parencite{Saugel2015}.

\begin{figure}
\centering
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_without_excl}
\caption{Basic four-quadrant plot.} \label{fig:trending_basic_4q}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_box}
\caption{Four-quadrant plot with rectangular exclusion area.}\label{fig:trending_basic_4q_excl_box}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_axis}
\caption{Four-quadrant plot with horizontal exclusion area.} \label{fig:trending_basic_4q_excl_axis}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_cross}
\caption{Four-quadrant plot with crossed exclusion area.}\label{fig:trending_basic_4q_excl_cross}
\end{subfigure}%
\caption{Illustrations of the four-quadrant plot with sample points and with and without exclusion areas. }
\label{fig:trending_4q}
\end{figure}
